---
title: "Practice / First Steps in Data Analytics"
author: Dr. Nishat Mohammad
date: 01/18/2024
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---



## Customer Transacation Data  
```{r Q1_explore_dataset}
# Question 1:  
#Load the data set from the link above into a data frame. Inspect the data set using the str() function. Each row represents a customer’s interactions with the #organization’s web store. The first column is the number of visits of a customer, the second the number of transactions of that customer, the third column is #the customer’s operating system, and the fourth column is the customer’s reported gender, while the last column is revenue, i.e., the total amount spent by #that customer.  
custxndata <- as.data.frame (read.csv("customertxndata.csv"))
str(custxndata)
# Get the Number of customers by using the number rows in the data
# I am not using the total gender (Male + Female) which would be 17,400 as the number of customers because some customer may not have provided their gender
customers <- nrow(custxndata)
data_vars <- colnames(custxndata)
```
The data  reveals `r paste(format(round(customers,0), big.mark = ",", scientific = FALSE))` customers with the following data: `r paste0('"',data_vars,'"')`


```{r Q2_summative_stats, echo = FALSE}
## Question 2:  
#In an R code chunk that is not displayed (i.e., set echo = F) Calculate the following summative statistics: total transaction amount (revenue), mean number #of visits, median revenue, standard deviation of revenue, most common gender. Exclude any cases where there is a missing value.  

summary(custxndata)

# Get the total transaction amount/Revenue
total_revenue <- sum(custxndata$rev, na.rm = T)

# Get the mean number of visits
Mean_visits <- mean(custxndata$numvisits, na.rm = T)

# Get the median Revenue
Median_revenue <- median(custxndata$rev, na.rm = T)

# Get the sd of Revenue
SD_revenue <- sd(custxndata$rev, na.rm = T)

# Get the most common gender
#help(unique)
#help(match)
valid_gender <- unique(custxndata$gender)
most_common_gender <- valid_gender[which.max(tabulate(match(custxndata$gender,valid_gender)))]

# Validate by using dplyr 
library(dplyr)
customer_gender <- custxndata%>%
  filter(!is.na(gender)) %>%
  group_by(gender) %>%
  count()
# Look at the customer_gender to check if Male is highest
customer_gender
total_customers_by_gender <- sum(customer_gender$n)
```
The summative statistics are shown in the summary above, the number of males and females for customers who shared their gender are also shown in the table. 

```{r Q3, echo = FALSE}
#Question 3:  
#Add a header to the R Notebook (## Data Analysis) and use the data calculated above to produce the narrative below, inserting the computed values in embedded #R code chunks – do not hard code the value as they might change if the data changes.

#There were 376 customers and the mean number of visits per customer was 3.4. The median revenue was US$ 35.8 (σ = 11.3). Most of the visitors were male.

#The numbers in the above text are not necessarily correct – they are added for illustration only. Also, make sure that the gender for the most visits is #calculated and not “hard coded”.  

```

## Data Analysis
There were `r paste(format(round(customers,0), big.mark = ",", scientific = FALSE))`  customers and the mean number of visits per customer was `r paste(format(round(Mean_visits,1), big.mark = ",", scientific = FALSE))`. The median revenue was `r paste("US$",format(round(Median_revenue,1), big.mark = ",", scientific = FALSE))` (`r paste("$\\sigma$ = ", format(round(SD_revenue, 1), big.mark = ",", scientific = FALSE))`). Most of the visitors were `r paste0('"',most_common_gender,'"')`.


### Visualizing Revenue Disparity in Gender  
```{r Q4_Bar_plot}
#Question 4:  
#Create a bar (aka column) chart of gender (x-axis) versus revenue (y-axis). Omit missing values, i.e., where gender is NA or missing. Use the plot() function # rather than functions from ggplot2. Show only the chart and not the code that generated it. Add markdown text to comment on what the chart means.  

Gender <- custxndata$gender
Revenue <- custxndata$rev
male_revenue <- sum(Revenue[which(Gender == "Male")])
female_revenue <- sum(Revenue[which(Gender == "Female")])

# Visualize with Barplot
barplot(c(male_revenue,female_revenue),
     main = "Bar Plot of Gender and Revenue",
     xlab = "Gender",ylab = "Revenue",
     names.arg = c("Female", "Male"),
     col = c("red","blue"))
```
The Bar Plot is titled Bar Plot of Gender and Revenue, with gender and revenue on the x and y axes respectively. males are in blue and females in red. The revenue of females is higher than males,female revenue is slightly above 500,000 and male revenue is close to 300,000.


### Correlation Data  
```{r Q5_Pearson_Correlation}
## Question 5:  
#What is the Pearson Moment of Correlation between number of visits and revenue? Comment on the correlation in your markdown, i.e., explain what it means and #on the overall strength and its relevance.  

Pcor_rev_visits <- cor(x=custxndata$numvisits, y=custxndata$rev, use = "complete.obs"
    , method = "pearson")
```
The Pearson correlation between number of visits and revenue is `r paste(format(round(Pcor_rev_visits,3), big.mark = ",", scientific = FALSE))`, It is positive and depicts that the two variables are directly proportional, increase in number of visits is commensurate with increase in revenue.


 

### Missing Data Highlights and Analysis:  
```{r Q6_missing_values}
## Question 6:  
#Which columns have missing data? How did you recognize them? How would you impute missing values? In your markdown, add comments on missing data and #imputation strategies. 

# Recognize the columns with NA using colSums()
total_na<- colSums(is.na(custxndata))
total_na
gender_na <- length(which(is.na(custxndata$gender)))
txnnum_na <- length(which(is.na(custxndata$numtxns)))
# Recognize the columns with NA using summary()
custxndata_summary <- within(custxndata, {
  gender <- factor(gender, labels = c("F", "M"))
  customer_os <- factor(OS, labels = c("Android", "iOS"))
})
summary(custxndata_summary)

# Impute NA using na.omit
valid_data <- na.omit(custxndata)
summary(valid_data)
```
The missing data (NA) can be recognized using the summary() function or using is.na() in the colSums() function.  
The total number of missing values  are `r paste(format(round(sum(total_na),0), big.mark = ",", scientific = FALSE))`.  
The Missing gender values are `r paste(format(round(gender_na,0), big.mark = ",", scientific = FALSE))`.  
The Missing number of transactions are `r paste(format(round(txnnum_na,0), big.mark = ",", scientific = FALSE))`.  


### Imputation of Missing Values:  
Imputation can be down using na.omit() or rm.na() and removal validated using is.na() again or summary() again. We can replace with the mean or mode for the transactions and assign a neutral gender to the missing gender values.  
```{r Q7_Impute_gender_transactions}
## Question 7:  
#Impute missing transaction and gender values. Use the mean for transaction (rounded to the nearest whole number) and the mode for gender. Recalculate the #descriptive statistics of (2) and repeat the markdown of (3) with the new (computed) values. Comment in the markdown on the difference.  

# Impute NA in transactions with mean value
mean_txn_col <- round(mean(custxndata$numtxns, na.rm = TRUE), digits = 0)
custxndata$numtxns[is.na(custxndata$numtxns)] <- mean_txn_col

# Impute NA in gender with mode
gender_mode <- max(custxndata$gender, na.rm = TRUE)
custxndata$gender[is.na(custxndata$gender)] <- gender_mode
```

```{r Q7_desc_stats}
#Calculating the descriptive statistics again
summary(custxndata)

# Get the total transaction amount/Revenue
new_total_revenue <- sum(custxndata$rev, na.rm = T)

# Get the mean number of visits
new_mean_visits <- mean(custxndata$numvisits, na.rm = T)

# Get the median Revenue
new_median_revenue <- median(custxndata$rev, na.rm = T)

# Get the sd of Revenue
new_sd_revenue <- sd(custxndata$rev, na.rm = T)

# Get the most common gender
help(unique)
help(match)
valid_gender <- unique(custxndata$gender)
new_most_common_gender <- valid_gender[which.max(tabulate(match(custxndata$gender,valid_gender)))]
new_most_common_gender

# Validate by using dplyr 
library(dplyr)
new_customer_gender <- custxndata%>%
  filter(!is.na(gender)) %>%
  group_by(gender) %>%
  count()
# Look at the customer_gender to check if Male is highest
new_customer_gender
```
### Data Analysis After Imputation  
There mean number of visits per customer was `r paste(format(round(new_mean_visits,1), big.mark = ",", scientific = FALSE))`.  
The median revenue was `r paste("US$",format(round(new_median_revenue,1), big.mark = ",", scientific = FALSE))` (`r paste("$\\sigma$ = ", format(round(new_sd_revenue, 1), big.mark = ",", scientific = FALSE))`).  
Most of the visitors were `r paste0('"',new_most_common_gender,'"')`.  




### Training and Validating sets:

```{r Q8_split_train_validate_dataset}
## Question 8:  
#Split the data set into two equally sized data sets where one can be used for training a model and the other for validation. Take every odd numbered case and #add them to the training data set and every even numbered case and add them to the validation data set, i.e., rows 1, 3, 5, 7, etc. are training data while #rows 2, 4, 6, etc. are validation data. Put this work into a new section with a new header and comments on the need for this.  


samp_cus_data <- custxndata %>%
  mutate(cus_data_split = seq_len(nrow(custxndata)) %% 2)

# Get rows in data to check the split
total_rows <- nrow(samp_cus_data)

# Train data set with odd case
training_dataset <- samp_cus_data %>%
  filter(cus_data_split == 1)
  
# Check split
Training_rows <- nrow(training_dataset)

# Validate data set with even case
validation_dataset <- samp_cus_data %>%
  filter(cus_data_split == 0)
# Check equal split
validation_rows <- nrow(validation_dataset)
```
#### Splitting Data  

The data set has been split into 2; training and validation sets. All the odd rows in the training set and all the even rows in the validation set. The values in the training set are `r paste(format(round(Training_rows,1), big.mark = ",", scientific = FALSE))` while the values in the validation set are `r paste(format(round(validation_rows,1), big.mark = ",", scientific = FALSE))`.  



#### Data Analysis After Split  
```{r Q9_mean_revenue_ of_sets}
## Question 9:  
#Calculate the mean revenue for the training and the validation data sets and compare them. Comment on the difference.  

# mean revenue training set
mean_rev_training <- mean(training_dataset$rev)

# mean revenue validation set
mean_rev_validation_df <- mean(validation_dataset$rev)
```

The mean revenue for the training set is `r paste("US$",format(round(mean_rev_training,2), big.mark = ",", scientific = FALSE))` and that for the validation set is `r paste("US$",format(round(mean_rev_validation_df,2), big.mark = ",", scientific = FALSE))`. The values are not so far apart, the validation set revenue mean is a bit higher than the training set revenue mean.



### Training, Testing and Validation sets  
```{r Q10_Final_Split}
## Question 10:  
#For many data mining and machine learning tasks, there are packages in R. Use the sample() function to split the data set, so that 60% is used for training #and 20% is used for testing, and another 20% is used for validation. To ensure that your code is reproducible and that everyone gets the same result, use the #number 77654 as your seed for the random number generator.  

final_split_data <- custxndata
set.seed(77654) 

samp <- sample(seq(1,3), size = nrow(final_split_data), replace = T, prob = c(0.6, 0.2, 0.2))

train_split <- final_split_data[samp == 1, ]
test_split <- final_split_data[samp == 2, ]
valid_split <- final_split_data[samp == 3,]

# Look at Sample distribution
traning_vals <- nrow(train_split)
testing_vals <- nrow(test_split)
validation_vals <- nrow(valid_split)
```
Splitted the data into three:  
The training data set with `r paste(format(round(traning_vals,0), big.mark = ",", scientific = FALSE))`.  
The testing data set with  `r paste(format(round(testing_vals,0), big.mark = ",", scientific = FALSE))`.  
The validation data set with `r paste(format(round(validation_vals,0), big.mark = ",", scientific = FALSE))`.  
